Three Problems for Sophistication

### 2015-01-20 Facticity/Facticity/paper.tex
...

\newcommand{\hide}[1]{}

@@

\section{Balancing under- and overfitting}

- define a default sophistication. PROBLEMS, PROBLEMS!
- ingewikkeld lemma, laat zien:
- can make it equal to C^K
- can make it bounded
- it's certainly not invariant
- any tweak that upsets balance -> can still go wrong the other way
- overfitting: one solution is to consider all representations that are within a constant of optimal, and then select the one with the smallest model. To effectively do this, the constant must depend on the numbering, but no-one knows how. In statistics overfitting is often addressed by imposing a penality on complex models. For sophistication a natural way of doing this is to allow non-prefix functions as models, which makes it comparatively cheap to put stuff in the data part. If you do this, overfitting cannot really happen anymore (?), but obviously only makes underfitting problem worse
- underfitting: most common solution is to restrict the model class. This helps (theorem), but obviously makes overfitting problem only worse
- only solution: upset balance both ways at the same time. There are approaches that do this and that therefore do not provably overfit or underfit, but (1) not clear if these methods are invariant (2) not clear if this sophistication still corresponds to original intuition: depth. See discussion

Overfitting: the singletons}
\label{section:balance}
We now come to the main issue plaguing sophistication. \sdr{Beetje bombastisch. En hierna zou ik zeggen, we first consider the simplest, most natural setting, \dots} If we use model class $\K$, the candidate set always contains two representations that do not provide a good separation. The first is a \emph{universal model}: a basic universal Turing machine. The second is a \emph{singleton model} a model which returns $x$ for any input. As we will show, in many settings, we can ensure that one of them is the \emph{only} model in the candidate set. The following lemma generalizes the principle:

\begin{restatable}{lemma}{coolone}
\label{lemma:thecoolone}
  Let $\psi$ be any acceptable enumeration of the partial recursive functions.
  Let $\M$ be any model class, let $\X$ be any set of binary sequences and let $D:\B\to\N$ be a computable decoding function with a prefix-free domain that maps function descriptions to their indices in $\psi$, i.e. if $f=\psi_{D(p)}$ then $p$ is a $D$-description of $f$. Let $\M'=\{\psi_i:i\in\tn{range}(D)\}$. Further assume there is a constant $c$ such that:\\
\-\hspace{1cm}(1)$\forall_{f\in\M'}:\min\{|p|:\psi_{D(p)}=f\}\le K^\psi(f)+c$\\
\-\hspace{1cm}(2)$\forall_{x\in\X}:L^{\M',\psi}(x)-L^{\M,\psi}(x)\le c$.\\
Then there is an enumeration $\phi$ of the partial recursive functions such that $S^\phi_{\M}(x) = |\bar 0|+S^\psi_{\M'}(x)$ for all $x\in\X$.
\end{restatable}
\noindent We can now show numberings for which the singletons are always selected:

\subsection(Overfitting}\label{section:overfitting}

\begin{restatable}[Overfitting]{lemma}{overfitting}
Let $\M \subseteq \K$ be a model class where for every $x\in\X$ there is a singleton model $f\in\M$ with $f(\epsilon)=x$. Then there is an enumeration $\phi$ of the prefix partial recursive functions, and a constant $c$, such that
\[
K(x)-S^{\M,\phi}(x)\le c
\]
for all $x\in\X$.
\end{restatable}
\noindent In short, this lemma tells us that there are numberings for which the singleton models always determine the sophistication. This means that there is a numbering $\phi$ such that the sophistication is equivalent to Kolmogorov complexity, which implies that this sophistication is unsound (by Lemma~\ref{lem:broken}).

\subsection{Underfitting: the universal model}
\label{section:underfitting}
On the other end of the spectrum, there exist representations with all information in the noise, using a universal model. We can easily show that if such models exist, we can choose our numbering so that they are always chosen.
\begin{restatable}[underfitting]{lemma}{underfitting}
Let $\M$ be a model class containing a universal model $u$, with the property that $\exists c \forall m \in \M : L^u(x) \leq L^m(x) + c \p$. Then, there exist numberings such that $\s^\M$ is bounded.
\end{restatable}
\noindent Most sophistication measures avoid the underfitting problem by limiting their model class, usually to either finite sets ($\F$) or total functions ($\T$ or $\P$). This eliminates the universal model, but as we shall see, it does not solve the problem for most data. The variants that are susceptible to the this type of underfitting are facticity \cite{adriaans2012facticity} and some versions of effective complexity \cite{gellmann1996information}. The latter aims to bypass the problem by selecting carefully from the candidate set, but as we have shown, for some numberings the candidate set contains only the universal model.
\sdr{Explain this is actually a widely recognised problem}

\subsection{The problem of depth}\label{section:depth}

We have seen that some variants of sophistication sidestep the simplest examples of underfitting by using total models. Does this ensure that these variants are not bounded? A well-known result in algorithmic statistics shows that there exist strings for which no finite set is as good a model as the singleton set. In this section we will focus on the variants $S^\T$ and $S^T_K$ \cite{vitanyi2004meaningful}, as those are the variants most resistant to proofs of non-invariance.
\sdr{Do I know this notation? Also, I wanna emphasize that the structure function underfits in this way.}
\sdr{Combine this lemma with the one after next}
\begin{restatable}{lemma}{nonstochastic}
There are infinitely  $x$ such that $S^\F_K(x) \geq K(x)/5$. \label{lemma:structure-function-is-not-bounded}
\end{restatable}

\sdr{Move this lemma to the appendix altogether}
\begin{restatable}{lemma}{vitanyi}
Let $f$ be a total, computable function, such that for some $d$, $f(d) = x$ and let $k = K(f) + |d|$. Then there exists a finite set $S$ containing $x$ and a constant $c$, such that $K(S) \leq K(f) + K(|d|) + c$ and $\log |S| \leq |d|$.\label{lemma:total-to-sets}
\end{restatable}
\noindent This lemma is a variation on \cite[Lemma~7.2]{vitanyi2004meaningful}. We use this lemma to transport the result of Theorem~\ref{lemma:structure-function-is-not-bounded} to the model class of total functions:

\begin{restatable}{theorem}{dogfood}
There are infinitely many $x$ with $S^\T_K(x) \geq K(x)/6$ and $S^\T(x) \geq K(x)/6$. 
\end{restatable} 
\noindent This tells us that $S^\T_K(x)$ and $S^\T(x)$ can truly grow to a significant part of the total description length. While this shows that sophistication, if defined with particular care, is not entirely vacuous, \sdr{hold on, it might still be completely meaningless and/or not invariant} we can shed some light on exactly which strings end up with high sophistication, using the principle of \emph{depth}. We will use the following definition:

@@

@@

\nonstochastic*

\begin{proof}
From \cite[Proposition~I.3 (b)]{gacs2001algorithmic} we know that for sufficiently large $i$ there are strings of length less than $i$ that are not $(i/4, i/4)$-stochastic. For each $i$, let $x_i$ be the smallest binary string that is not $(i/4, i/4)$-stochastic. Since every string is $(\alpha,\alpha)$-stochastic for sufficiently large $\alpha$, this sequence contains infinitely many distinct binary strings. Now pick any $i$ that is large enough that (a) $|x_i|<i$ and (b) $K(x_i)/5 < i/4$. By definition of $(\alpha,\beta)$-stochasticity, if a string is not $(i/4,i/4)$-stochastic, then for every finite set $S$ containing $x$, either (a) $\log|S|\ge K(x_i)+i/4$, in which case $S$ cannot be a candidate representation for $x$ in $C^\F_K$, or (b) $K(S)>i/4>K(x_i)/5$. This proves the theorem.
\end{proof}

\vitanyi*
\begin{proof}
Let $S = f\left(\{0,1\}^{|d|}\right)$. Since $f$ is total, this set can be explicity computed from a description of $f$ and a description of $|d|$, which tells us that there is a constant $c$ such that $K(S) \leq K(f) + K(|d|) + c$.

Since $S$ is the image of a set of size $2^{|d|}$ under $f$, we have $\log |S| \leq |d|$.
\end{proof}

\dogfood*
\begin{proof}
We continue from the proof of Theorem~\ref{lemma:structure-function-is-not-bounded}. Suppose towards contradiction that either $C^\T_K(x_i)$ or $C^\T_{\tau^\circ}(x_i)$ contains a candidate $f$ with $K(f)\le i/5$. Then by Lemma~\ref{lemma:total-to-sets}, there is a set $S$ with $\log|S|\le|d|\le\tau^\circ(x)\le K(x)$, so (a) in the proof of Theorem~\ref{lemma:structure-function-is-not-bounded} above cannot be the case, and $K(S)\le K(f)+K(|d|)+c\le i/5+K(|d|)+c$, which contradicts (b) in the same proof for large enough $i$. Therefore such candidates $f$ cannot exist and all candidates must satisfy $K(f)>i/5>K(x_i)/6$.
\end{proof}

>>>>

\newcommand{\hide}[1]{}
\newcommand{\old}[1]{}

@@

\section{Balancing under- and overfitting}
\label{section:balance}

- define a default sophistication. PROBLEMS, PROBLEMS!

\begin{restatable}{lemma}{coolone}
\label{lemma:thecoolone}
  Let $\psi$ be any acceptable enumeration of the partial recursive functions.
  Let $\M$ be any model class, let $\X$ be any set of binary sequences and let $D:\B\to\N$ be a computable decoding function with a prefix-free domain that maps function descriptions to their indices in $\psi$, i.e. if $f=\psi_{D(p)}$ then $p$ is a $D$-description of $f$. Let $\M'=\{\psi_i:i\in\tn{range}(D)\}$. Further assume there is a constant $c$ such that:\\
\-\hspace{1cm}(1)$\forall_{f\in\M'}:\min\{|p|:\psi_{D(p)}=f\}\le K^\psi(f)+c$\\
\-\hspace{1cm}(2)$\forall_{x\in\X}:L^{\M',\psi}(x)-L^{\M,\psi}(x)\le c$.\\
Then there is an enumeration $\phi$ of the partial recursive functions such that $S^\phi_{\M}(x) = |\bar 0|+S^\psi_{\M'}(x)$ for all $x\in\X$.
\end{restatable}

- ingewikkeld lemma, laat zien:\\
- can make it equal to $C^K$
\begin{restatable}[Overfitting]{lemma}{overfitting}
Let $\M \subseteq \K$ be a model class where for every $x\in\X$ there is a singleton model $f\in\M$ with $f(\epsilon)=x$. Then there is an enumeration $\phi$ of the prefix partial recursive functions, and a constant $c$, such that
\[
K(x)-S^{\M,\phi}(x)\le c
\]
for all $x\in\X$.
\end{restatable}

- can make it bounded
\begin{restatable}[underfitting]{lemma}{underfitting}
Let $\M$ be a model class containing a universal model $u$, with the property that $\exists c \forall m \in \M : L^u(x) \leq L^m(x) + c \p$. Then, there exist numberings such that $\s^\M$ is bounded.
\end{restatable}

- it's certainly not invariant \\
- any tweak that upsets balance -> can still go wrong the other way

- overfitting: one solution is to consider all representations that are within a constant of optimal, and then select the one with the smallest model. To effectively do this, the constant must depend on the numbering, but no-one knows how. \\
- In statistics overfitting is often addressed by imposing a penalty on complex models. For sophistication a natural way of doing this is to allow non-prefix functions as models, which makes it comparatively cheap to put stuff in the data part. If you do this, overfitting cannot really happen anymore (?), but obviously only makes underfitting problem worse

\subsection{Underfitting}
- underfitting: most common solution is to restrict the model class. This helps

\begin{restatable}{theorem}{dogfood}
There are infinitely many $x$ with $\s^{\F,\phi,c}(x) \geq C^{\K}(x)/5$ and $\s^{\F,\phi,c}(x) \geq C^{\K}(x)/6$. 
\end{restatable}

(theorem), but obviously makes overfitting problem only worse

- only solution: upset balance both ways at the same time. There are approaches that do this and that therefore do not provably overfit or underfit, but (1) not clear if these methods are invariant (2) not clear if this sophistication still corresponds to original intuition: depth. See discussion

@@

\old{

We now come to the main issue plaguing sophistication. \sdr{Beetje bombastisch. En hierna zou ik zeggen, we first consider the simplest, most natural setting, \dots} If we use model class $\K$, the candidate set always contains two representations that do not provide a good separation. The first is a \emph{universal model}: a basic universal Turing machine. The second is a \emph{singleton model} a model which returns $x$ for any input. As we will show, in many settings, we can ensure that one of them is the \emph{only} model in the candidate set. The following lemma generalizes the principle:

\noindent We can now show numberings for which the singletons are always selected:

\noindent In short, this lemma tells us that there are numberings for which the singleton models always determine the sophistication. This means that there is a numbering $\phi$ such that the sophistication is equivalent to Kolmogorov complexity, which implies that this sophistication is unsound (by Lemma~\ref{lem:broken}).

On the other end of the spectrum, there exist representations with all information in the noise, using a universal model. We can easily show that if such models exist, we can choose our numbering so that they are always chosen.

\noindent Most sophistication measures avoid the underfitting problem by limiting their model class, usually to either finite sets ($\F$) or total functions ($\T$ or $\P$). This eliminates the universal model, but as we shall see, it does not solve the problem for most data. The variants that are susceptible to the this type of underfitting are facticity \cite{adriaans2012facticity} and some versions of effective complexity \cite{gellmann1996information}. The latter aims to bypass the problem by selecting carefully from the candidate set, but as we have shown, for some numberings the candidate set contains only the universal model.
\sdr{Explain this is actually a widely recognised problem}

\noindent This tells us that $S^\T_K(x)$ and $S^\T(x)$ can truly grow to a significant part of the total description length. While this shows that sophistication, if defined with particular care, is not entirely vacuous, \sdr{hold on, it might still be completely meaningless and/or not invariant} we can shed some light on exactly which strings end up with high sophistication, using the principle of \emph{depth}. We will use the following definition:

}

@@

We first show that representions using $\T$ can be reduced to representations in $\F$ with a small overhead.
\pb{Definitie nonstochastic strings uit \cite[Proposition~I.3 (b)]{gacs2001algorithmic}}

\begin{restatable}{lemma}{vitanyi}
Let $f$ be a total, computable function, such that for some $d$, $f(d) = x$ and let $k = K(f) + |d|$. Then there exists a finite set $S$ containing $x$ and a constant $c$, such that $K(S) \leq K(f) + K(|d|) + c$ and $\log |S| \leq |d|$.\label{lemma:total-to-sets}
\end{restatable}
\begin{proof}
Let $S = f\left(\{0,1\}^{|d|}\right)$. Since $f$ is total, this set can be explicity computed from a description of $f$ and a description of $|d|$, which tells us that there is a constant $c$ such that $K(S) \leq K(f) + K(|d|) + c$.

Since $S$ is the image of a set of size $2^{|d|}$ under $f$, we have $\log |S| \leq |d|$.
\end{proof}
\noindent This lemma is a variation on \cite[Lemma~7.2]{vitanyi2004meaningful}.

We use this to show that sophistications with a restricted model class are not bounded:
\dogfood*
\begin{proof}
From \cite[Proposition~I.3 (b)]{gacs2001algorithmic} we know that for sufficiently large $i$ there are strings of length less than $i$ that are not $(i/4, i/4)$-stochastic. For each $i$, let $x_i$ be the smallest binary string that is not $(i/4, i/4)$-stochastic. Since every string is $(\alpha,\alpha)$-stochastic for sufficiently large $\alpha$, this sequence contains infinitely many distinct binary strings. Now pick any $i$ that is large enough that (a) $|x_i|<i$ and (b) $C^\K(x_i)/5 < i/4$. By definition of $(\alpha,\beta)$-stochasticity, if a string is not $(i/4,i/4)$-stochastic, then for every finite set $S$ containing $x$, either (a) $\log|S|\ge C^\K(x_i)+i/4$, in which case $S$ cannot be a candidate representation for $x$ in \pb{candidate set}, or (b) $C^\K(S)>i/4>C^\K(x_i)/5$. This proves the first inequality.

Now, suppose towards contradiction that \pb{$\T$ candidate set} contains a candidate $f$ with $C^\K(f)\le i/5$. Then by Lemma~\ref{lemma:total-to-sets}, there is a set $S$ with $\log|S|\le|d|\le\tau^\circ(x)\le C^\K(x)$, so (a) above cannot be the case, and $K(S)\le K(f)+K(|d|)+c\le i/5+K(|d|)+c$, which contradicts (b) for large enough $i$. Therefore such candidates $f$ cannot exist and all candidates must satisfy $K(f)>i/5>K(x_i)/6$.
\end{proof}
